QPLIB: A Library of Quadratic Programming Instances

### 2017-02-07 QPLIB-3.tex
more accurate description of the advantages of .qplib provided

In this section we present all the steps performed in order to build the
new library. In \S\ref{subsec:instColl}, we describe the set of gathered
instances. In \S\ref{subsec:selection} we present the features used to
classify the instances and we discuss the issues concerning the format of
the instances. Finally, in \S\ref{subsec:final set}, we describe the
selection process used to filter the instances and we graphically present
the main features of the selected instances.

@@

At the end of this process we gathered more than eight thousand
instances. Three fourths of them contained discrete variables, while
the remaining ones contained only continuous variables. More in details,
we gathered $\approx 1800$ Quadratic Binary Linear (QBL) instances,
$\approx 2000$ Quadratic Continuous Quadratic (QCQ) instances, and
and $\approx 2500$ Quadratic General Quadratic (QGQ) instances. We
also gathered $\approx 1000$ Convex General Convex (CGC) instances. We
gathered relatively fewer Quadratic Binary Quadratic (QBQ), Convex
Continuous Convex (CCC) and Convex Mixed Convex (CMC) instances, i.e.,
$\approx 150$ instances, $\approx 200$ instances and $\approx 200$
instances respectively. Finally, we gathered only $17$ Quadratic Mixed
Linear (QML) instances. In the call for instances, no specific formats
requirements were imposed for the submissions.

To evaluate the instances we decided, for practical reasons, to use
GAMS as common platform for all the experiments involving commercial and
non-commercial solvers. For this reason, we decided to translate all
instances into the GAMS format (\texttt{.gms}).
% In a preliminary phase, all the instances received were divided
%according to their format and subsequently translated.
%In \S\ref{subsec:tools}
%%~\ref{subsec:CP_convex}
%the tools used
%to translate an instance from a given format to the \texttt{.gms} format
%%to translate an instance from and to a given format to the
%\texttt{.gms} format
%are described in more detail.\\
%
In addition, we have introduced a new, specific \texttt{.qplib} format. This
new format is capable of describing all the instances of the library in a sparse form.
In comparison to a more \emph{high level} and human readable format like \texttt{.gms}, the new
format presents three advantages: it is easier to read by a self-made parser, it produces smaller files, and the order of variables and constraints is fixed. See the \S\ref{sec:format} for more details.

@@

For each instance of the starting set, we collected important characteristics
which allowed us to classify the instances into the QP categories described in
\S\ref{sec:QPbasic}. As far as the variable types are concerned, we
collected the following information:
%
\begin{itemize}
 \item number of binary variables; % (\emph{\# bin}),
 %
 \item number of integer variables; % (\emph{\# int}),
 %
 \item number of continuous variables. % (\emph{\# cont}).
\end{itemize}
%
If at least one binary or integer variable is present, then the instance is
categorized as \emph{discrete}, otherwise it is categorized as \emph{continuous}.
As far as the objective function is concerned, we gathered the following
information:
%
\begin{itemize}
 \item percentage of positive and negative eigenvalues of the Hessian $Q^0$;
       % (\emph{\% neg eig}),
 %
 \item density of the Hessian $Q^0$ (number of nonzero entries divided by the total
       number of entries). %(\emph{\% dens}).
\end{itemize}
%

The number of positive (i.e., larger than $10^{-16}$) and negative (i.e., smaller than $-10^{-16}$) eigenvalues of $Q^0$ allows us to identify the
objective function type, as in presence of at least one negative (positive) eigenvalue
the objective function is nonconvex (nonconcave). Finally, as far as the constraint types
are concerned, we collected the following information:
%
\begin{itemize}
 \item number of linear constraints, % (\emph{\# lin}),
 %
 \item number of quadratic constraints, % (\emph{\# quad}),
 %
 \item number of convex constraints, % (\emph{\# conv}),
 %
 \item number of variable bounds (for non-binary variables). %(\emph{\# box}).
\end{itemize}
%
A constraint is considered quadratic if it contains at least one nonzero in
the quadratic term. Among the quadratic constraints, the ones with only
non-negative eigenvalues are classified as convex constraints.
Note that this might lead to classify some instances that have conic constraints as non-convex ones, although their feasible region is in fact convex. However, only some solvers are capable of properly exploiting this property.
 All this information allowed us to analyse the gathered instances and to perform the
filters described in the the next paragraph.

@@

During the development of the library, a discussion ensued about
the expected goals that we wanted to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item represent as much as possible all the different categories of QP
       problems;
 %
 \item gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item include for each of the categories a set of well-diversified
       instances;
 %
 \item obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve the aforementioned goals, we performed the following two
filters, applied in cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S\ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in GAMS, which number depends on the category
       of the instance under consideration, cf.~Table \ref{t:solvers}.
       We then filtered according to a first
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analysed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives (e.g., very similar size, same donor,\ldots). Finally, in order to only
       keep computationally challenging instances we ran a a complete solver for QGQ with a time limit of 120 seconds; all the
       instances which have been solved to proven optimality within this time limit
       were discarded.
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $251$ discrete instances and
$116$ continuous instances.

@@

We now analyse the features of the instances selected to be part of the
library. The characteristics of the instances are presented in Table
\ref{tab:DD} for \emph{discrete} instances (*\{B,M,I,G\}*) and in Table
\ref{tab:CC} for \emph{continuous} ones (*C*). For each category, the tables
report in column ``$\#$'' the corresponding number of instances. It can be seen
that the final set well respects the original distribution of the gathered
instances among the different categories. Indeed, the discrete categories
(LMQ) or (QBL) are well represented by $118$ and $59$ instances,
respectively. Similarly, the continuous categories (LCQ) and (QCQ) are well
represented by $50$ and $17$ instances, respectively. Moreover, the library
actually covers the large majority of all possible categories of instances.

@@

Speaking of nonconvexity, Figure~\ref{fig:pic_neg_eig} (left) focuses on
the instances with a quadratic objective function, ordered by the
the percentage ``hard'' eigenvalues in the Hessian $Q^0$ (vertical axis), by which
we mean negative eigenvalues in case of a minimization problem and positive eigenvalues in case of a maximization problem.
The instances are mostly clustered around two values. About 25\% of the instances have a convex (if minimization) or concave (if maximization) objective function,
i.e., they have 0\% of ``hard'' eigenvalues. Among the others, a vast
majority has around 50\% of ``hard'' eigenvalues. However, instances
with high or low percentages of ``hard'' eigenvalues are present, too.

@@

Additional details on the instance features can be found in \S\ref{sec:instance_details}.

@@

\subsection{Website}
%Select subset or categories of instances (EXTRACT FROM THE LIBRARY A SUBSET OF INSTANCES WITH SPECIFIC CHARACTERISTICS)

The instances of QPLIB are publicly accessible at the website \url{http://qplib.zib.de}, which was created by extending
scripts and tools initially developed for MINLPLib\,2~\cite{Vigerske2014}.
%
Beyond all instances in GAMS, \texttt{.lp}, AMPL (\texttt{.mod}), and \texttt{.qplib} format, the website provides a rich set of metadata for each
instance: the three letter problem classification (as described in Section \ref{subsec:final set}), basic properties such as the number of variables and constraints of
different types, the sense and convexity/concavity of the objective function, and information on the nonzero structure of the
problem.
%
In addition, we display a visualization of the sparsity patterns of the Jacobian and the Hessian matrix of the
Lagrangian function.  In the plots of the Jacobian nonzero pattern, the linear and nonlinear entries are distinguished
by color.  Figure \ref{fig:sparsitypattern} shows an example for instance QPLIB\_2967.

>>>>

In this section we present all the steps we performed in order to build
the new instance library. In \S\ref{subsec:instColl}, we describe the set
of gathered instances, and
in \S\ref{subsec:selection} we present the features used to
classify the instances.
%and discuss issues concerning the format of the instances.
We describe the selection process used to filter the instances, and
graphically present the main features of the selected instances in
\S\ref{subsec:final set}, while in \S\ref{subsec:website}
we provide information on how to access the test collection.

@@

At the end of this process we had gathered more than eight thousand
instances. Three quarters of them contained discrete variables, while
the remainder contained only continuous variables. In more detail,
we gathered $\approx 1800$ Quadratic Binary Linear (QBL) instances,
$\approx 2000$ Quadratic Continuous Quadratic (QCQ) instances, and
and $\approx 2500$ Quadratic General Quadratic (QGQ) instances. We
also received $\approx 1000$ Convex General Convex (CGC) instances. We
obtained relatively fewer Quadratic Binary Quadratic (QBQ), Convex
Continuous Convex (CCC) and Convex Mixed Convex (CMC) instances,
($\approx 150$, $\approx 200$ and $\approx 200$ instances respectively).
Finally, we found only $17$ Quadratic Mixed
Linear (QML) instances. In the call for instances, no specific formats
requirements were imposed for the submissions.

To evaluate the instances we decided, for practical reasons, to use
GAMS as common platform for all our final selection computations.
%involving commercial and non-commercial solvers.
For this reason, we translated all the
instances we received into the GAMS format (\texttt{.gms}).
% In a preliminary phase, all the instances received were divided
%according to their format and subsequently translated.
%In \S\ref{subsec:tools}
%%~\ref{subsec:CP_convex}
%the tools used
%to translate an instance from a given format to the \texttt{.gms} format
%%to translate an instance from and to a given format to the
%\texttt{.gms} format
%are described in more detail.\\
%

@@

For each instance in this large starting set,
we collected important characteristics
which allowed us to classify the instances into the QP categories described in
\S\ref{sec:QPbasic}. As far as the variable types are concerned, we
collected the following information:
%
\begin{itemize}
 \item the number of binary variables; % (\emph{\# bin}),
 %
 \item the number of integer variables; and % (\emph{\# int}),
 %
 \item the number of continuous variables. % (\emph{\# cont}).
\end{itemize}
%
If at least one binary or integer variable is present, then the instance is
categorized as \emph{discrete}, otherwise it is categorized as \emph{continuous}.
As far as the objective function is concerned, we gathered the following
information:
%
\begin{itemize}
 \item the percentage of positive and negative eigenvalues of the Hessian
      $Q^0$; and
       % (\emph{\% neg eig}), and
 %
 \item the density of the Hessian $Q^0$ (number of nonzero entries divided by the total
       number of entries). %(\emph{\% dens}).
\end{itemize}
%

The number of positive (i.e., larger than $10^{-16}$) and negative (i.e., smaller than $-10^{-16}$) eigenvalues of $Q^0$ allowed us to identify the
objective function type, as in presence of at least one negative (positive) eigenvalue
the objective function is nonconvex (nonconcave). Finally, as far as the constraint types
are concerned, we collected the following information:
%
\begin{itemize}
 \item the number of linear constraints, % (\emph{\# lin}),
 %
 \item the number of quadratic constraints, % (\emph{\# quad}),
 %
 \item the number of convex constraints, and % (\emph{\# conv}),
 %
 \item the number of variable bounds (for non-binary variables). %(\emph{\# box}).
\end{itemize}
%
A constraint is considered quadratic if it contains at least one nonzero in
a quadratic term (if present). Among the quadratic constraints, the ones whose
Hessians have only non-negative eigenvalues (when $c_u^i < \infty$) and
and non-positive eigenvalues (when $c_l^i > - \infty$)
are classified as convex constraints; thus, a quadratic constraint with
two sided, finite bounds is non-convex.
Note that this might might occasionally lead us to classify some instances that have conic constraints as non-convex ones, although their feasible region is in fact convex---fortunately, only some solvers are capable of properly exploiting this property.
 All this information allowed us to analyse the gathered instances and to perform the
filters described in the the next paragraph.

@@

During the development of the library, a discussion ensued about
the expected goals that we wished to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item to represent as far as possible all the different categories of QP
       problems;
 %
 \item to gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item to include, for each of the categories, a set of well-diversified
       instances; and
 %
 \item to obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve such goals, we performed the following two filters, applied in
a cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S\ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in GAMS, which number depends on the category
       of the instance under consideration, cf.~Table \ref{t:solvers}.
       We then filtered according to a first
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analysed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives (e.g., very similar size, same donor,\ldots). Finally, in order to only
       keep computationally challenging instances we ran a a complete solver for QGQ with a time limit of 120 seconds; all the
       instances which have been solved to proven optimality within this time limit
       were discarded.
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $251$ discrete instances and
$116$ continuous instances.

@@

We now analyse the features of the instances selected to be part of the
instance library. The characteristics of the instances are presented in Table
\ref{tab:DD} for \emph{discrete} instances
(\textit{*}\{\textit{B,M,I,G}\}\textit{*}) and in Table
\ref{tab:CC} for \emph{continuous} ones (\textit{*C*}).
For each category, the tables
report in column ``$\#$'' the corresponding number of instances. It can be seen
that the final set well respects the original distribution of the gathered
instances among the different categories. Indeed, the discrete categories
(\textit{LMQ}) or (\textit{QBL}) are well represented by $118$ and $59$
instances, respectively. Similarly, the continuous categories
(\textit{LCQ}) and (\textit{QCQ}) are well
represented by $50$ and $17$ instances, respectively. Moreover, the library
actually covers the large majority of all possible categories of instances.

@@

On the theme of nonconvexity, Figure~\ref{fig:pic_neg_eig} (left) focuses on
the instances with a quadratic objective function, ordered by the
the percentage ``hard'' eigenvalues in the Hessian $Q^0$ (vertical axis), by which
we mean negative eigenvalues in case of a minimization problem and positive eigenvalues in case of a maximization problem.
The instances are mostly clustered around two values. About 25\% of the instances have a convex (if minimization) or concave (if maximization) objective function,
i.e., they have 0\% of ``hard'' eigenvalues. Among the others, a vast
majority has around 50\% of ``hard'' eigenvalues. However, instances
with high or low percentages of ``hard'' eigenvalues are present, too.

@@

Additional details on the instance features can be found in Appendix~\ref{sec:instance_details}.

@@

\subsection{Website}\label{subsec:website}
%Select subset or categories of instances (EXTRACT FROM THE LIBRARY A SUBSET OF INSTANCES WITH SPECIFIC CHARACTERISTICS)

The QPLIB instances are publicly accessible at the website \url{http://qplib.zib.de}, which was created by extending
scripts and tools initially developed for MINLPLib\,2~\cite{Vigerske2014}.
%
We provide all instances in GAMS (\texttt{.gms}), AMPL (\texttt{.mod}), CPLEX (\texttt{.lp})
\cite{,cplex}, and  QPLIB (\texttt{.qplib}) formats. The latter is a new format
specifically for QP instances. In comparison to more \emph{high level}
formats such as \texttt{.gms} and \texttt{.lp}, the new
format offers three main advantages: it is easier to read by a stand-alone
parser (provided), it typically produces smaller files, and it permits the
inclusion of two-sided inequalities without needless repetition of data.
%the order of variables and constraints is fixed.
See Appendix~\ref{sec:format} for more details.

Beyond the example instances, the website provides a rich set of metadata for
each instance: the three letter problem classification (as described in \S\ref{subsec:final set}), basic properties such as the number of variables and constraints of
different types, the sense and convexity/concavity of the objective function, and information on the nonzero structure of the
problem.
%
In addition, we display a visualization of the sparsity patterns of the Jacobian and the Hessian matrix of the
Lagrangian function.  In the plots of the Jacobian nonzero pattern, the linear and nonlinear entries are distinguished
by color.  Figure \ref{fig:sparsitypattern} shows an example for instance QPLIB\_2967.

### 2017-02-07 conclusion.tex
more accurate description of the advantages of .qplib provided

 This manuscript describes the first comprehensive library of instances for Quadratic Programming (QP). Since QP comprises  different and ``varied'' categories of problems, we propose a classification and we briefly discuss the main classes of solution methods for QP.

We then describe the steps of the adopted process used to filter the gathered instances  in order to build the new library. The goals we purposed are of different natures and we tried to build a library which is computationally challenging and as broad as possible, i.e., it represents the largest possible categories of QP problems, while remaining of manageable size.

>>>>

 This manuscript describes the first comprehensive library of instances for Quadratic Programming (QP). Since QP comprises  different and ``varied'' categories of problems, we proposed a classification and we briefly discuss the main classes of solution methods for QP.

We then describe the steps of the adopted process used to filter the gathered instances  in order to build the new library. Our design goals were to build a library which is computationally challenging and as broad as possible, i.e., it represents the largest possible categories of QP problems, while remaining of manageable size. We have also proposed a stand-alone QP format that is intended
for the convenient exchange and use of our QP instances.

### 2017-02-07 methods.tex
Update methods.tex
Added citations to BARON-related algorithmic papers
more accurate description of the advantages of .qplib provided

The solvers in Table \ref{t:solvers} which implement asymptotically complete methods are \oqnlp, \msnlp, \knitro, and certain sub-solvers of \lgo. Specifically, we consider the following methods:
%
\begin{itemize}
 \item global adaptive random search (\lgo);
 %
 \item multi-start (\knitro, \lgo, \msnlp, \oqnlp); specifically, the former three apply to \textit{QCQ} whereas the latter to \textit{QGQ}.
\end{itemize}

@@

Typically, bounds in the optimization direction are computed by means of convex relaxations \cite{mccormick,BeLeLiMaWa08}, which replace nonconvex terms $t(x)$ with linearization variables $\hat{t}$, and then replace the corresponding defining constraints $\hat{t} = t(x)$ by means of lower and upper (respectively, convex and concave) bounding functions $\hat{t} \ge \underline{t}(x)$ and $\hat{t} \le\bar{t}(x)$. This is actually where finite (and tight) bounds on the variables are crucial, which differentiate also in practice the bounded case from the unbounded one. Different strategies are used when the nonconvexities are only quadratic \cite{MiFl13,bliek}.

>>>>

The solvers in Table \ref{t:solvers} which implement asymptotically complete methods are \oqnlp, \msnlp, \knitro, and certain sub-solvers of \lgo. Specifically, we consider the following methods:
%
\begin{itemize}
 \item global adaptive random search (\lgo);
 %
 \item multi-start (\knitro, \lgo, \msnlp, \oqnlp); specifically, the first three apply to \textit{QCQ} and the last to \textit{QGQ}.
\end{itemize}

@@

Typically, bounds in the optimization direction are computed by means of convex relaxations \cite{mccormick,BeLeLiMaWa08}, which replace nonconvex terms $t(x)$ with linearization variables $\hat{t}$, and then replace the corresponding defining constraints $\hat{t} = t(x)$ by means of lower and upper (respectively, convex and concave) bounding functions $\hat{t} \ge \underline{t}(x)$ and $\hat{t} \le\bar{t}(x)$. This is actually where finite (and tight) bounds on the variables are crucial, which differentiate also in practice the bounded case from the unbounded one. Different strategies are used when the nonconvexities are only quadratic \cite{bst:09:oms,MiFl13,zs:13:oms,zs:13:iecr,bliek}.
